Challenging Fuel Cycle Modeling Assumptions: Facility and Time Step Discretization Effects

### 2016-02-05 paper.tex
remove improper usage of 'due to'

\myabstract{
Due to the diversity of fuel cycle simulator modeling assumptions, direct
comparison and benchmarking can be difficult.  In 2012 the Organisation for
Economic Co-operation and Development (OECD) completed a benchmark study that
is perhaps the most complete published comparison performed.  Despite this,
various results from the simulators were often significantly different due to
inconsistencies in modeling decisions involving reprocessing strategies,
refueling behavior, reactor end-of-life handling, etc.  This work identifies
and quantifies the effects of selected modeling choices that may sometimes be
taken for granted in the fuel cycle simulation domain.  Four scenarios are
compared using combinations of either fleet-based or individually modeled
reactors with either monthly or quarterly (3-month) time steps.  The scenarios
approximate a transition from the current U.S. once-through light water
reactor (LWR) fleet to a full sodium fast reactor (SFR) fuel cycle. The Cyclus
fuel cycle simulator's plug-in facility capability along with its market-like
dynamic material routing allow it to be used as a level playing field for
comparing the scenarios.  When under supply-constraint pressure, the four
cases exhibit noticeably different behavior.  Fleet-based modeling is more
efficient in supply-constrained environments at the expense of losing insight
on issues such as realistically suboptimal fuel distribution and challenges in
reactor refueling cycle staggering.  Finer-grained time steps also enable more
efficient material use in supply-constrained environments resulting in much lower
standing inventories of separated Pu.  Large simulations with fleet-based
reactors run much more quickly than their individual reactor counterparts.
Gaining a better understanding of how these and other modeling choices affect
fuel cycle dynamics will enable making more deliberate decisions with respect
to trade-offs such as computational investment vs.  realism.
}

@@

The diversity of assumptions embedded within various fuel cycle simulators pose challenges for direct
comparison and benchmarking.  In 2012 the Organisation for
Economic Co-operation and Development (OECD) completed a benchmark study
\cite{oecd2012benchmark} that is perhaps one of the most complete published
comparisons performed.  Despite this, however, various results from the
different simulators were often significantly different due to inconsistent
modeling decisions involving reprocessing strategies, refueling behavior,
reactor end-of-life handling, etc.  This work attempts to identify and
quantify the effects of selected modeling choices relative to facility modeling
discretization (i.e. individual facilities vs. aggregate fleets) and time step
duration.  A better understanding of how these affect fuel cycle dynamics will
allow for making more deliberate decisions with respect to trade-offs such as
computational investment vs. realism.

@@

Traditionally, many fuel cycle simulators have used a system dynamics approach
\cite{forrester_industrial_1961} to modeling. These simulators model several
\emph{stocks} representing isotopes or isotope groups that have quantities that change
over time due to \emph{flows} between them.  The flows are determined by equations
that are a function of the state of the system (i.e. the system's stocks).  A
constructed system is then solved in discrete or continuous time depending on
the limitations of the software and constraints imposed by the model.  In
system dynamics based simulators, the number of stocks and potential flows between
them is generally a static property of the modeled system, making it somewhat
difficult to swap different reactor models in and out of the same scenario
without wiring them into the simulator manually.

@@

The parameters for each case were also selected to keep capacity factors for
individual reactors very roughly around 90\%, although they are actually
closer to 80\% due to integer variable constraints on the case MI and QI
reactor configuration.

@@

Quantifying this effect directly is difficult.  One way of observing
it is to compare fleet reactor scenarios that have different time step
durations (i.e. cases MF and QF).  Because the fleet reactors have perfect fuel
sharing (described below) and no noise from refueling outages, differences observed will
partly be due to quantized shutdown.

@@

Figure \ref{fig:power} shows an overall view of the generated power over time
for each of the four cases.  Figure \ref{fig:power-rel} normalizes the Figure
\ref{fig:power} curves to the expected exponential power curve magnifying some
interesting differences between the four cases.  For the first 100 years, the
four cases behave somewhat similarly, although the individually modeled
reactors in cases MI and QI actually go offline and back online for refueling
causing more variance.  Around year 100, a fuel shortage begins and more
significant differences between the four cases become apparent. This fuel
shortage lasts until about year 140. The power generated in case QF is slightly
lower than in case MF during the fuel shortage years. This is partly due to
the \emph{quantized shutdown effect}.  The longer 3-month time step in case QF 
results in reactor capacity going offline longer than necessary.

@@

\begin{figure*}[!h]
    \centering
    \includegraphics[width=1.0\textwidth]{exp2/power-rel.eps}
    \caption[Normalized power]{
        Generated power is normalized to the expected 1\% exponential growth
        curve for all four cases.  A fuel shortage occurs from about year 100
        to about year 140.  The \emph{cycle staggering effect} can be seen as
        excessive divergence above and below $1.0$ in cases MI and QI where
        the shortage increases cycle synchronization significantly. The
        \emph{quantized shutdown effect} causes part of the discrepancy
        between quarterly and monthly time steps for fleet reactor modeling
        (cases MF and QF) during the shortage; a longer time step causes more
        reactor outage than necessary. The consistent jumping up and down for
        the fleet reactor cases (even pre-shortage) is due to the misalignment of reactors going
        offline (multiples of 12 months) and the build period (21 months).
    }
    \label{fig:power-rel}
\end{figure*}

@@

\begin{figure}[!h]
    \centering
    \includegraphics[width=1.0\columnwidth]{exp2/unfueled.eps}
    \caption[Cumulative offline power due to fuel shortage]{
        Cumulative offline power for reactors that had delayed cycle start due
        to fuel shortage. The difference between individual and fleet based
        modeling is primarily a result of the \emph{fuel sharing effect}. The
        difference between monthly and quarterly time steps is primarily
        caused by the \emph{inventory drawdown effect}.
    }
    \label{fig:unfueled}
\end{figure}

Quantifying imperfect fuel sharing is a bit difficult, but Figure
\ref{fig:badshare} provides one way to see the \emph{fuel sharing effect}
showing wasted batch-months computed using Equation \ref{eq:badshare}.  Fleet
reactor modeling, by design, exhibits perfect fuel sharing (zero
inefficiency) and so is not included in this figure.  Every batch that is
given to a reactor that ends up not being able to start its cycle at the
scheduled time (due to fuel shortage) adds one to this cumulative total for each
time step the cycle start is delayed.

@@

\begin{figure}[!h]
    \centering
    \includegraphics[width=1.0\columnwidth]{exp2/badshare.eps}
    \caption[Cumulative unnecessary idling fuel]{
        Fuel sharing inefficiency approximated by the cumulative number of
        fuel batches given to each reactor on each refueling period multiplied
        by the number of time steps that reactor had to delay the start of its
        next cycle (due to fuel shortage).  Fleet reactors implicitly have perfect fuel
        sharing and so are not included here. The large difference
        between cases MI and QI is mostly due to the \emph{inventory drawdown
        effect}. 
    }
    \label{fig:badshare}
\end{figure}

@@

The large differences seen between monthly and quarterly time step cases in
Figures \ref{fig:unfueled} and \ref{fig:badshare} are primarily caused by the
\emph{inventory drawdown effect}.  Figure \ref{fig:puinv} helps to visualize
this effect.  The black dots in the figure represent impulse flows of
available Pu from fuel fabrication and into fast reactors.  Blue lines represent
the flows into Pu inventory available for making fuel. The in-flows are not
shown for the individual reactor cases because they are somewhat messy and
obscure other information in the plots. The other colored curves show Pu inventory
available for fabricating fast reactor fuel.  When the inventory curves meet and
force down the out-flows, fuel shortages occur.  While this is most clear for
cases MF and QF, the individual reactor scenarios (cases MI and QI) also have many
points during the shortage where the outflow point lies exactly on top of the
inventory point --- indicating that all available inventory is being
transferred.

@@

Throughout most of the shortage, the aggregate outflow of Pu from separated
inventory is roughly the same in cases MF and QF.  However, due to the longer
time steps in case QF, the shortage starts slightly sooner because Pu inventory
must be maintained at a higher limit as described above.  The small difference
in outage start time has feedback effects.  Not only are extra reactors
offline, but those extra reactors are also discharging less aggregate spent
fuel for recycling.  This increases the discrepancy between the two
cases. Eventually the shortage in case MF ends slightly before it does in case
QF (shown most clearly in Figure \ref{fig:puinv-compare-zoom}). This feedback effect occurs because the reactors have a
conversion ratio greater than 1.0.

@@

Figure \ref{fig:puinv-compare} shows the same inventory curves from Figure
\ref{fig:puinv} superimposed. Perhaps counter-intuitively, lower separated
plutonium inventory during the shortage generally indicates fewer unfueled
reactors.  Lower inventory levels mean available Pu is being utilized more
efficiently rather than sitting idle.  Cases QI and QF with their longer time
steps show higher inventories indicating a more severe shortage.  Cases QI and
QF take longer to recover from the shortage and fall a bit behind with respect
to building up Pu stocks due to the delayed contributions of more unfueled
reactors to the separated Pu pool.  This can be seen most clearly in Figure
\ref{fig:puinv-compare-zoom}.  Cases MI and QF recover Pu inventory
post-shortage at about the same time and rate --- this is consistent with their
approximately equivalent cumulative offline power capacity curves in Figure
\ref{fig:unfueled}.

>>>>

\myabstract{
Due to the diversity of fuel cycle simulator modeling assumptions, direct
comparison and benchmarking can be difficult.  In 2012 the Organisation for
Economic Co-operation and Development (OECD) completed a benchmark study that
is perhaps the most complete published comparison performed.  Despite this,
various results from the simulators were often significantly different because of
inconsistencies in modeling decisions involving reprocessing strategies,
refueling behavior, reactor end-of-life handling, etc.  This work identifies
and quantifies the effects of selected modeling choices that may sometimes be
taken for granted in the fuel cycle simulation domain.  Four scenarios are
compared using combinations of either fleet-based or individually modeled
reactors with either monthly or quarterly (3-month) time steps.  The scenarios
approximate a transition from the current U.S. once-through light water
reactor (LWR) fleet to a full sodium fast reactor (SFR) fuel cycle. The Cyclus
fuel cycle simulator's plug-in facility capability along with its market-like
dynamic material routing allow it to be used as a level playing field for
comparing the scenarios.  When under supply-constraint pressure, the four
cases exhibit noticeably different behavior.  Fleet-based modeling is more
efficient in supply-constrained environments at the expense of losing insight
on issues such as realistically suboptimal fuel distribution and challenges in
reactor refueling cycle staggering.  Finer-grained time steps also enable more
efficient material use in supply-constrained environments resulting in much lower
standing inventories of separated Pu.  Large simulations with fleet-based
reactors run much more quickly than their individual reactor counterparts.
Gaining a better understanding of how these and other modeling choices affect
fuel cycle dynamics will enable making more deliberate decisions with respect
to trade-offs such as computational investment vs.  realism.
}

@@

The diversity of assumptions embedded within various fuel cycle simulators pose challenges for direct
comparison and benchmarking.  In 2012 the Organisation for
Economic Co-operation and Development (OECD) completed a benchmark study
\cite{oecd2012benchmark} that is perhaps one of the most complete published
comparisons performed.  Despite this, however, various results from the
different simulators were often significantly different because of inconsistent
modeling decisions involving reprocessing strategies, refueling behavior,
reactor end-of-life handling, etc.  This work attempts to identify and
quantify the effects of selected modeling choices relative to facility modeling
discretization (i.e. individual facilities vs. aggregate fleets) and time step
duration.  A better understanding of how these affect fuel cycle dynamics will
allow for making more deliberate decisions with respect to trade-offs such as
computational investment vs. realism.

@@

Traditionally, many fuel cycle simulators have used a system dynamics approach
\cite{forrester_industrial_1961} to modeling. These simulators model several
\emph{stocks} representing isotopes or isotope groups that have quantities that change
over time because of \emph{flows} between them.  The flows are determined by equations
that are a function of the state of the system (i.e. the system's stocks).  A
constructed system is then solved in discrete or continuous time depending on
the limitations of the software and constraints imposed by the model.  In
system dynamics based simulators, the number of stocks and potential flows between
them is generally a static property of the modeled system, making it somewhat
difficult to swap different reactor models in and out of the same scenario
without wiring them into the simulator manually.

@@

The parameters for each case were also selected to keep capacity factors for
individual reactors very roughly around 90\%, although they are actually
closer to 80\% because of integer variable constraints on the case MI and QI
reactor configuration.

@@

Quantifying this effect directly is difficult.  One way of observing
it is to compare fleet reactor scenarios that have different time step
durations (i.e. cases MF and QF).  Because the fleet reactors have perfect fuel
sharing (described below) and no noise from refueling outages, differences observed will
partly be a result of quantized shutdown.

@@

Figure \ref{fig:power} shows an overall view of the generated power over time
for each of the four cases.  Figure \ref{fig:power-rel} normalizes the Figure
\ref{fig:power} curves to the expected exponential power curve magnifying some
interesting differences between the four cases.  For the first 100 years, the
four cases behave somewhat similarly, although the individually modeled
reactors in cases MI and QI actually go offline and back online for refueling
causing more variance.  Around year 100, a fuel shortage begins and more
significant differences between the four cases become apparent. This fuel
shortage lasts until about year 140. The power generated in case QF is slightly
lower than in case MF during the fuel shortage years. This is partly a result of
the \emph{quantized shutdown effect}.  The longer 3-month time step in case QF 
results in reactor capacity going offline longer than necessary.

@@

\begin{figure*}[!h]
    \centering
    \includegraphics[width=1.0\textwidth]{exp2/power-rel.eps}
    \caption[Normalized power]{
        Generated power is normalized to the expected 1\% exponential growth
        curve for all four cases.  A fuel shortage occurs from about year 100
        to about year 140.  The \emph{cycle staggering effect} can be seen as
        excessive divergence above and below $1.0$ in cases MI and QI where
        the shortage increases cycle synchronization significantly. The
        \emph{quantized shutdown effect} causes part of the discrepancy
        between quarterly and monthly time steps for fleet reactor modeling
        (cases MF and QF) during the shortage; a longer time step causes more
        reactor outage than necessary. The consistent jumping up and down for
        the fleet reactor cases (even pre-shortage) is caused by the misalignment of reactors going
        offline (multiples of 12 months) and the build period (21 months).
    }
    \label{fig:power-rel}
\end{figure*}

@@

\begin{figure}[!h]
    \centering
    \includegraphics[width=1.0\columnwidth]{exp2/unfueled.eps}
    \caption[Cumulative offline power caused by fuel shortage]{
        Cumulative offline power for reactors that had delayed cycle start caused
        by fuel shortage. The difference between individual and fleet based
        modeling is primarily a result of the \emph{fuel sharing effect}. The
        difference between monthly and quarterly time steps is primarily
        a result of the \emph{inventory drawdown effect}.
    }
    \label{fig:unfueled}
\end{figure}

Quantifying imperfect fuel sharing is a bit difficult, but Figure
\ref{fig:badshare} provides one way to see the \emph{fuel sharing effect}
showing wasted batch-months computed using Equation \ref{eq:badshare}.  Fleet
reactor modeling, by design, exhibits perfect fuel sharing (zero
inefficiency) and so is not included in this figure.  Every batch that is
given to a reactor that ends up not being able to start its cycle at the
scheduled time (caused by fuel shortage) adds one to this cumulative total for each
time step the cycle start is delayed.

@@

\begin{figure}[!h]
    \centering
    \includegraphics[width=1.0\columnwidth]{exp2/badshare.eps}
    \caption[Cumulative unnecessary idling fuel]{
        Fuel sharing inefficiency approximated by the cumulative number of
        fuel batches given to each reactor on each refueling period multiplied
        by the number of time steps that reactor had to delay the start of its
        next cycle (caused by fuel shortage).  Fleet reactors implicitly have perfect fuel
        sharing and so are not included here. The large difference
        between cases MI and QI is mostly a result of the \emph{inventory drawdown
        effect}. 
    }
    \label{fig:badshare}
\end{figure}

@@

The large differences seen between monthly and quarterly time step cases in
Figures \ref{fig:unfueled} and \ref{fig:badshare} are primarily a result of the
\emph{inventory drawdown effect}.  Figure \ref{fig:puinv} helps to visualize
this effect.  The black dots in the figure represent impulse flows of
available Pu from fuel fabrication and into fast reactors.  Blue lines represent
the flows into Pu inventory available for making fuel. The in-flows are not
shown for the individual reactor cases because they are somewhat messy and
obscure other information in the plots. The other colored curves show Pu inventory
available for fabricating fast reactor fuel.  When the inventory curves meet and
force down the out-flows, fuel shortages occur.  While this is most clear for
cases MF and QF, the individual reactor scenarios (cases MI and QI) also have many
points during the shortage where the outflow point lies exactly on top of the
inventory point --- indicating that all available inventory is being
transferred.

@@

Throughout most of the shortage, the aggregate outflow of Pu from separated
inventory is roughly the same in cases MF and QF.  However, because of the longer
time steps in case QF, the shortage starts slightly sooner because Pu inventory
must be maintained at a higher limit as described above.  The small difference
in outage start time has feedback effects.  Not only are extra reactors
offline, but those extra reactors are also discharging less aggregate spent
fuel for recycling.  This increases the discrepancy between the two
cases. Eventually the shortage in case MF ends slightly before it does in case
QF (shown most clearly in Figure \ref{fig:puinv-compare-zoom}). This feedback effect occurs because the reactors have a
conversion ratio greater than 1.0.

@@

Figure \ref{fig:puinv-compare} shows the same inventory curves from Figure
\ref{fig:puinv} superimposed. Perhaps counter-intuitively, lower separated
plutonium inventory during the shortage generally indicates fewer unfueled
reactors.  Lower inventory levels mean available Pu is being utilized more
efficiently rather than sitting idle.  Cases QI and QF with their longer time
steps show higher inventories indicating a more severe shortage.  Cases QI and
QF take longer to recover from the shortage and fall a bit behind with respect
to building up Pu stocks because of the delayed contributions of more unfueled
reactors to the separated Pu pool.  This can be seen most clearly in Figure
\ref{fig:puinv-compare-zoom}.  Cases MI and QF recover Pu inventory
post-shortage at about the same time and rate --- this is consistent with their
approximately equivalent cumulative offline power capacity curves in Figure
\ref{fig:unfueled}.
